---
title: "phenotypes"
output: pdf_document
date: "2023-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cleaning Phenotypes
### K. Uckele October 10, 2023

```{r libraries, echo=FALSE}
library(googlesheets4)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(EnvStats)
library(stringr)
```

# Red area (nectar guide data)
## Load and filter data
```{r load red area}
redarea <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="red_area")

#make a new column that makes a unique ID for each plant
redarea <- mutate(redarea, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
redarea <- redarea %>% dplyr::select(-date, -photo_set, -plant_type, -ID, -labellum_photo, -stamen_photo)

#rename columns
redarea <- redarea %>% dplyr::rename("RALA" = "red_labellum",
                                   "RAST" = "red_stamen")

#take a quick look at data frame
redarea
```
## Collapse the replicate observations
```{r red area collapse reps}
#First, make a function to calculate the mode of categorical variables
#this function will output NA if there is a tie
Modes <- function(x) {
  ux <- unique(na.omit(x))
  tab <- tabulate(match(x, ux))
  if(sum(tab == max(tab)) == 2) {NA} 
  else {ux[tab == max(tab)]}
}

#collapse the replicates by taking means
#first take the mean of continuous data
redarea_by_ID <- redarea %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers red area}
## Boxplot of red labellum
ggplot(redarea_by_ID) +
  aes(x = "", y = RALA) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
## Identify potential outliers using the IQR criterion
# print outlier values
sort(boxplot.stats(redarea_by_ID$RALA)$out, decreasing = TRUE)
#Rosner test
rosnerTest(redarea_by_ID$RALA, k=length(boxplot.stats(redarea_by_ID$RALA)$out))
#no outliers detected

## Boxplot of red stamen
ggplot(redarea_by_ID) +
  aes(x = "", y = RAST) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
boxplot.stats(redarea_by_ID$RAST)$out
#Rosner test
rosnerTest(redarea_by_ID$RAST, k=length(boxplot.stats(redarea_by_ID$RAST)$out))
#two outliers detected
## Remove outliers
redarea_by_ID <- redarea_by_ID %>%
  mutate(RAST = na_if(RAST, 43.800)) %>%
  mutate(RAST = na_if(RAST, 19.933))
```

# Inflorescence-level data
## Load and filter data
```{r load inflorescence}
#get access to google sheets
gs4_auth(email = "kuckele@ucsc.edu")
#read in the tab of interest and convert it to a commonly named df
inflor <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="inflorescences")

#make a new column that makes a unique ID for each plant
inflor <- mutate(inflor, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
inflor <- inflor %>% dplyr::select(-date, -rep, -plant_type, -ID, -typeID)

#make sure categorical variables are factors
inflor$visible_EFnectar <- as.factor(inflor$visible_EFnectar)
inflor$visible_guides <- as.factor(inflor$visible_guides)

#rename columns
inflor <- inflor %>% dplyr::rename("INFA" = "infl_angle",
                                   "VEFN" = "visible_EFnectar",
                                   "CAL" = "callus_length",
                                   "VNG" = "visible_guides")

#take a quick look at data frame
inflor
```
## Collapse the replicate observations
```{r inflor collapse reps}
#collapse the replicates by taking means and modes
#take the mean of continuous data and the mode of the factors and characters
inflor_by_ID <- inflor %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Fill in VNG based on red area data
```{r impute VNG}
# Join the two data frames by the unique_ID column
inflor_by_ID <- inflor_by_ID %>%
  left_join(redarea_by_ID, by = "unique_ID") %>%
  mutate(VNG = case_when(
    is.na(VNG) & (RAST > 0 | RALA > 0) ~ factor(1, levels = c(0, 1)),
    is.na(VNG) & RAST == 0 & RALA == 0 ~ factor(0, levels = c(0, 1)),
    TRUE ~ VNG  # Keep the original value if the condition is not met
  )) %>%
  # Remove the columns from redarea_by_ID if no longer needed
  select(-RAST, -RALA)
```

## Identify potential outliers
```{r inflor outliers}
## Boxplot of inflorescence angle
ggplot(inflor_by_ID) +
  aes(x = "", y = INFA) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values based on IQR criterion
sort(boxplot.stats(inflor_by_ID$INFA)$out, decreasing = TRUE)
#Rosner test
rosnerTest(inflor_by_ID$INFA, k=length(boxplot.stats(inflor_by_ID$INFA)$out))
#no outliers detected based on Rosner test

## Boxplot of callus length
ggplot(inflor_by_ID) +
  aes(x = "", y = CAL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values based on IQR criterion
boxplot.stats(inflor_by_ID$CAL)$out
#Rosner test
rosnerTest(inflor_by_ID$CAL, k=length(boxplot.stats(inflor_by_ID$CAL)$out))
#no outliers detected based on Rosner test
```

# Flower morphology data
## Load and filter data
```{r load flower morph}
flo_morph <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="flower_morphology")

#make a new column that makes a unique ID for each plant
flo_morph <- mutate(flo_morph, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
flo_morph <- flo_morph %>% dplyr::select(-date, -plant_type, -ID, -rep)

#rename columns
flo_morph <- flo_morph %>% dplyr::rename("COL" = "Corolla_Length",
                                   "COLL" = "Corolla_Lobe_Length",
                                   "STAE" = "Stamen_exsertion",
                                   "TUA" = "tube_angle",
                                   "STATL" = "stamen_tip",
                                   "LABL" = "Labellum_Length",
                                   "LABW" = "Labellum_Width",
                                   "CLL" = "Labellum_lobe",
                                   "STAL" = "Stamen_Length",
                                   "STAW" = "Stamen_width",
                                   "ANL" = "Anther_Length",
                                   "ANW" = "Anther_width",
                                   "STIW" = "Stigma_Width",
                                   "STYL" = "Style_length")

#quick look at data
flo_morph
```
## Collapse the replicate observations
```{r flower morph collapse reps}
flo_morph_by_ID <- flo_morph %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers flower morph}
## Corolla Length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = COL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$COL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$COL, k=length(boxplot.stats(flo_morph_by_ID$COL)$out))
#one outlier detected
## Remove outliers
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(COL = na_if(COL, 27.2))

## Corolla_Lobe_Length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = COLL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$COLL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$COLL, k=length(boxplot.stats(flo_morph_by_ID$COLL)$out))
#no outliers detected

## Stamen_exsertion
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STAE) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STAE)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$STAE, k=length(boxplot.stats(flo_morph_by_ID$STAE)$out))
#no outliers detected

## tube_angle
ggplot(flo_morph_by_ID) +
  aes(x = "", y = TUA) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$TUA)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$TUA, k=length(boxplot.stats(flo_morph_by_ID$TUA)$out))
#three outliers detected
## Remove outliers
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(TUA = round(TUA, digits=0)) %>%
  mutate(TUA = na_if(TUA, 43)) %>%
  mutate(TUA = na_if(TUA, 0))

## stamen_tip
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STATL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STATL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$STATL, k=length(boxplot.stats(flo_morph_by_ID$STATL)$out))
#no outliers detected

## Labellum_Length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = LABL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$LABL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$LABL, k=length(boxplot.stats(flo_morph_by_ID$LABL)$out))
#no outliers detected

## Labellum_Width
ggplot(flo_morph_by_ID) +
  aes(x = "", y = LABW) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# no outliers detected

## Labellum_lobe
ggplot(flo_morph_by_ID) +
  aes(x = "", y = CLL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$CLL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$CLL, k=length(boxplot.stats(flo_morph_by_ID$CLL)$out))
#one outlier detected
## Remove outliers
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(CLL = na_if(CLL, 12.4)) 

## Stamen_Length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STAL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STAL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$STAL, k=length(boxplot.stats(flo_morph_by_ID$STAL)$out))
#no outliers detected

## Stamen_width
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STAW) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STAW)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$STAW, k=length(boxplot.stats(flo_morph_by_ID$STAW)$out))
#one outlier detected
## Remove outlier
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(STAW = na_if(STAW, 5.5)) 
 
## Anther_Length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = ANL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$ANL)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$ANL, k=length(boxplot.stats(flo_morph_by_ID$ANL)$out))
#one outlier detected
## Remove outliers
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(ANL = na_if(ANL, 9.15)) 

## Anther_width
ggplot(flo_morph_by_ID) +
  aes(x = "", y = ANW) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$ANW)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$ANW, k=1)
#no outliers detected

## Stigma_Width
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STIW) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STIW)$out, decreasing = TRUE)
#Rosner test
rosnerTest(flo_morph_by_ID$STIW, k=length(boxplot.stats(flo_morph_by_ID$STIW)$out))
#one outlier detected
## Remove outliers
flo_morph_by_ID <- flo_morph_by_ID %>%
  mutate(STIW = na_if(STIW, 4.6)) 

## Style_length
ggplot(flo_morph_by_ID) +
  aes(x = "", y = STYL) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(flo_morph_by_ID$STYL)$out, decreasing = TRUE)
# no outliers detected
```

# Nectar data
## Load and filter data
```{r load nectar}
nectar <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="nectar")

#make a new column that makes a unique ID for each plant
nectar <- mutate(nectar, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
nectar <- nectar %>% dplyr::select(-year, -date, -plant_type, -ID, -rep, -fl_capsize)

# EFN_percent_sucrose is dependent on EFN_H2Ovolume_uL
# Split EFN_percent_sucrose into two different variables: 1) diluted in 40 uL or
# 2) diluted in 50 uL (because these cannot be directly compared)

nectar <- nectar %>%
  mutate(
    EFN_percent_sucrose_H2Ovolume_40uL = if_else(
      EFN_H2Ovolume_uL == 40,       # Condition: Check if H2O volume is 40 µL
      EFN_percent_sucrose,          # If TRUE: Assign the value of EFN_percent_sucrose
      NA_real_                      # If FALSE: Assign NA (as a numeric NA)
    ), 
    EFN_percent_sucrose_H2Ovolume_50uL = if_else(
      EFN_H2Ovolume_uL == 50,       # Condition: Check if H2O volume is 50 µL
      EFN_percent_sucrose,          # If TRUE: Assign the value of EFN_percent_sucrose
      NA_real_  
    )
  )

#exclude columns we don't need
# I'm removing EFN_percent_sucrose_H2Ovolume_50uL because there are not enough observations to QTL map
nectar <- nectar %>% dplyr::select(-fl_nectarlength_mm, -fl_mg_sucrose, -EFN_H2Ovolume_uL, -EFN_percent_sucrose, -EFN_mg_sucrose, -EFN_percent_sucrose_H2Ovolume_50uL)

#rename columns
nectar <- nectar %>% dplyr::rename("VFN" = "fl_nectarvolume_uL", # volume floral nectar
                                   "FNSC" = "fl_percent_sucrose", # percent sucrose floral nectar
                                   "EFNSC40" = "EFN_percent_sucrose_H2Ovolume_40uL") # percent sucrose (diluted in 40 uL) EFN 

#quick look at data
nectar
```

## Collapse the replicate observations
```{r nectar collapse reps}
nectar_by_ID <- nectar %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers nectar}
## fl_nectarlength_mm
ggplot(nectar_by_ID) +
  aes(x = "", y = VFN) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(nectar_by_ID$VFN)$out, decreasing = TRUE)
#Rosner test
rosnerTest(nectar_by_ID$VFN, k=length(boxplot.stats(nectar_by_ID$VFN)$out))
#no outliers detected

## fl_percent_sucrose
ggplot(nectar_by_ID) +
  aes(x = "", y = FNSC) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(nectar_by_ID$FNSC)$out, decreasing = TRUE)
#Rosner test
rosnerTest(nectar_by_ID$FNSC, k=length(boxplot.stats(nectar_by_ID$FNSC)$out))
#three outliers detected
## Remove outliers
nectar_by_ID <- nectar_by_ID %>%
  mutate(FNSC = na_if(FNSC, 0.0)) %>%
  mutate(FNSC = na_if(FNSC, 8.5)) %>%
  mutate(FNSC = na_if(FNSC, 12.5))

## Extrafloral nectar percent sucrose (diluted in 40 uL)
ggplot(nectar_by_ID) +
  aes(x = "", y = EFNSC40) +
  geom_boxplot(fill = "#0c4c8a") +
  theme_minimal()
# print outlier values
sort(boxplot.stats(nectar_by_ID$EFNSC40)$out, decreasing = TRUE)
#Rosner test
rosnerTest(nectar_by_ID$EFNSC40, k=length(boxplot.stats(nectar_by_ID$EFNSC40)$out))
#two outliers detected
## Remove outliers
nectar_by_ID <- nectar_by_ID %>%
  mutate(EFNSC40 = round(EFNSC40, digits=2)) %>%
  mutate(EFNSC40 = na_if(EFNSC40, 37.00000)) 
## make sure outliers are removed
rosnerTest(nectar_by_ID$EFNSC40, k=2)
```

# Combine and write data
```{r combine data}
data <- inflor_by_ID %>% 
  full_join(redarea_by_ID, by="unique_ID") %>%
  full_join(flo_morph_by_ID, by="unique_ID") %>%
  full_join(nectar_by_ID, by="unique_ID")


write.csv(data, "~/Dropbox/Costus/costus-genetic-mapping/phenotype/results/processed_data/phenotypic_data_no_outliers_with_parents_F1s.csv", quote = FALSE, row.names = FALSE)
```

## Exclude parents and F1s
```{r exclude parents and F1s}
#exclude parents and F1s
inflor_by_ID_filtered <- inflor_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

redarea_by_ID_filtered <- redarea_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

flo_morph_by_ID_filtered <- flo_morph_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

nectar_by_ID_filtered <- nectar_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

data <- inflor_by_ID_filtered %>% 
  full_join(redarea_by_ID_filtered, by="unique_ID") %>%
  full_join(flo_morph_by_ID_filtered, by="unique_ID") %>%
  full_join(nectar_by_ID_filtered, by="unique_ID")


write.csv(data, "~/Dropbox/Costus/costus-genetic-mapping/phenotype/results/processed_data/phenotypic_data_no_outliers_exclude_parents_F1s.csv", quote = FALSE, row.names = FALSE)

```
