---
title: "phenotypes"
output: pdf_document
date: "2023-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Cleaning Phenotypes
### K. Uckele October 10, 2023

```{r libraries, echo=FALSE}
library(googlesheets4)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(EnvStats)
library(stringr)
library(patchwork)  
```

# Red area (nectar guide data)
## Load and filter data
```{r load red area}
redarea <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="red_area")

#make a new column that makes a unique ID for each plant
redarea <- mutate(redarea, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
redarea <- redarea %>% dplyr::select(-date, -photo_set, -plant_type, -ID, -labellum_photo, -stamen_photo)

#rename columns
redarea <- redarea %>% dplyr::rename("RALA" = "red_labellum",
                                   "RAST" = "red_stamen")

#take a quick look at data frame
redarea
```
## Collapse the replicate observations
```{r red area collapse reps}
#First, make a function to calculate the mode of categorical variables
#this function will output NA if there is a tie
Modes <- function(x) {
  ux <- unique(na.omit(x))
  tab <- tabulate(match(x, ux))
  if(sum(tab == max(tab)) == 2) {NA} 
  else {ux[tab == max(tab)]}
}

#collapse the replicates by taking means
#first take the mean of continuous data
redarea_by_ID <- redarea %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers red area}

traits <- setdiff(names(redarea_by_ID), "unique_ID")

result <- reduce(traits,           # loop over traits
  .init = list(data = redarea_by_ID, plots = list()),
  .f = function(acc, tr) {

    dat <- acc$data
    vec <- dat[[tr]]

    ## 1. Rosner test – how many to remove?
    k <- rosnerTest(vec, k = length(boxplot.stats(vec)$out))$n.outliers

    ## 2. pick k single most‑extreme values
    outs    <- boxplot.stats(vec)$out
    center  <- median(vec, na.rm = TRUE)
    extreme <- outs[order(abs(outs - center), decreasing = TRUE)][seq_len(k)]
    
    message(sprintf("Trait %s – %d extreme values flagged (unique_IDs starting with 'P' and 'F1' kept)",
                  tr, length(extreme)))

    ## 3. save BEFORE plot
    p_before <- ggplot(dat, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: before"))

    ## 4. remove those extremes unless ID starts with P or F1
    dat2 <- dat %>%
      mutate(
        {{ tr }} := if_else(
          .data[[tr]] %in% extreme & !grepl("^(P|F1)", unique_ID),
          NA_real_,
          .data[[tr]]
        )
      )

    ## 5. AFTER plot
    p_after <- ggplot(dat2, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: after"))

    ## 6. sanity‑check: Rosner should find zero now
    if (k > 0) {
      remaining <- rosnerTest(dat2[[tr]],
                            k = length(extreme))$n.outliers
      if (remaining > 0)
      warning(glue::glue("{tr}: {remaining} outlier(s) still present"))
    }
    
    ## 7. store updated data and plots
    acc$data  <- dat2
    acc$plots[[tr]] <- p_before | p_after
    acc
  })

redarea_by_ID <- result$data
plots    <- result$plots          # list of before|after panels

# combine all plots into grid
wrap_plots(plots, nrow = length(traits))   # grid of all traits
```

# Inflorescence-level data
## Load and filter data
```{r load inflorescence}
#get access to google sheets
gs4_auth(email = "kuckele@ucsc.edu")

inflor <- read_sheet(
  ss    = "19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs",
  sheet = "inflorescences",
  col_types = "??????l?l"
) %>% 
  # add your extra wrangling steps
  mutate(unique_ID = paste0(plant_type, "_", ID)) %>% 
  select(-date, -rep, -plant_type, -ID, -typeID) %>% 
  # convert the two visibility flags to factors 
  #mutate(across(c(visible_EFnectar, visible_guides), as.factor)) %>%
  dplyr::rename("INFA" = "infl_angle",
                "VEFN" = "visible_EFnectar",
                "CAL" = "callus_length",
                "VNG" = "visible_guides")

#take a quick look at data frame
inflor
```
## Collapse the replicate observations
```{r inflor collapse reps}
#collapse the replicates by taking means and modes
#take the mean of continuous data and the mode of the factors and characters
inflor_by_ID <- inflor %>% 
  group_by(unique_ID) %>% 
  summarise(
    across(where(is.double),   ~ mean(.x, na.rm = TRUE)),  # numeric columns
    across(where(is.character), Modes),                    # your custom mode fn
    across(where(is.logical), Modes),                      # your custom mode fn
    .groups = "drop"                                       # ungroups automatically
  )
```

## Fill in VNG based on red area data
```{r impute VNG}
# Join the two data frames by the unique_ID column
inflor_by_ID <- inflor_by_ID %>%
  left_join(redarea_by_ID, by = "unique_ID") %>%
  mutate(VNG = case_when(
    is.na(VNG) & (RAST > 0 | RALA > 0) ~ TRUE,
    is.na(VNG) & RAST == 0 & RALA == 0 ~ FALSE,
    TRUE ~ VNG  # Keep the original value if the condition is not met
  )) %>%
  # Remove the columns from redarea_by_ID if no longer needed
  select(-RAST, -RALA)
```

## Identify potential outliers
```{r inflor outliers}

traits <- setdiff(names(inflor_by_ID), c("unique_ID", "VEFN", "VNG"))

result <- reduce(traits,           # loop over traits
  .init = list(data = inflor_by_ID, plots = list()),
  .f = function(acc, tr) {

    dat <- acc$data
    vec <- dat[[tr]]

    ## 1. Rosner test – how many to remove?
    k <- rosnerTest(vec, k = length(boxplot.stats(vec)$out))$n.outliers

    ## 2. pick k single most‑extreme values
    outs    <- boxplot.stats(vec)$out
    center  <- median(vec, na.rm = TRUE)
    extreme <- outs[order(abs(outs - center), decreasing = TRUE)][seq_len(k)]
    
    message(sprintf("\n\nTrait %s – %d extreme values flagged (unique_IDs starting with 'P' and 'F1' kept)",
                  tr, length(extreme)))

    ## 3. save BEFORE plot
    p_before <- ggplot(dat, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: before"))

    ## 4. remove those extremes unless ID starts with P or F1
    dat2 <- dat %>%
      mutate(
        {{ tr }} := if_else(
          .data[[tr]] %in% extreme & !grepl("^(P|F1)", unique_ID),
          NA_real_,
          .data[[tr]]
        )
      )

    ## 5. AFTER plot
    p_after <- ggplot(dat2, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: after"))

    ## 6. sanity‑check: Rosner should find zero now
    if (k > 0) {
      
      # remove the parent and F1 observations from the data before testing
      vec_keep <- dat %>%     
      filter(!grepl("^(P|F1)", unique_ID)) %>%
      pull({{ tr }})
      
      remaining <- rosnerTest(vec_keep,
                            k = length(extreme))$n.outliers
      if (remaining > 0)
      warning(glue::glue("{tr}: {remaining} outlier(s) still present"))
    }
    
    ## 7. store updated data and plots
    acc$data  <- dat2
    acc$plots[[tr]] <- p_before | p_after
    acc
  })

inflor_by_ID <- result$data
plots    <- result$plots          # list of before|after panels

# combine all plots into grid
wrap_plots(plots, nrow = length(traits))   # grid of all traits
```

# Flower morphology data
## Load and filter data
```{r load flower morph}
flo_morph <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="flower_morphology")

#make a new column that makes a unique ID for each plant
flo_morph <- mutate(flo_morph, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
flo_morph <- flo_morph %>% dplyr::select(-date, -plant_type, -ID, -rep)

#rename columns
flo_morph <- flo_morph %>% dplyr::rename("COL" = "Corolla_Length",
                                   "COLL" = "Corolla_Lobe_Length",
                                   "STAE" = "Stamen_exsertion",
                                   "TUA" = "tube_angle",
                                   "STATL" = "stamen_tip",
                                   "LABL" = "Labellum_Length",
                                   "LABW" = "Labellum_Width",
                                   "CLL" = "Labellum_lobe",
                                   "STAL" = "Stamen_Length",
                                   "STAW" = "Stamen_width",
                                   "ANL" = "Anther_Length",
                                   "ANW" = "Anther_width",
                                   "STIW" = "Stigma_Width",
                                   "STYL" = "Style_length")

#quick look at data
flo_morph
```
## Collapse the replicate observations
```{r flower morph collapse reps}
flo_morph_by_ID <- flo_morph %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers flower morph}

traits <- setdiff(names(flo_morph_by_ID), c("unique_ID"))

result <- reduce(traits,           # loop over traits
  .init = list(data = flo_morph_by_ID, plots = list()),
  .f = function(acc, tr) {

    dat <- acc$data
    
    #vec <- dat[[tr]]
    
    vec <- dat %>%                       # rows that CAN be tested/removed
      filter(!grepl("^(P|F1)", unique_ID)) %>%
      pull({{ tr }})

    ## 1. Rosner test – how many to remove?
    if (length(boxplot.stats(vec)$out) == 0) {
      test.k <- 1
    } else {
      test.k <- length(boxplot.stats(vec)$out)
    }
    
    k <- rosnerTest(vec, k = test.k)$n.outliers

    ## 2. pick k single most‑extreme values
    outs    <- boxplot.stats(vec)$out
    center  <- median(vec, na.rm = TRUE)
    extreme <- outs[order(abs(outs - center), decreasing = TRUE)][seq_len(k)]
    
    message(sprintf("\n\nTrait %s – %d extreme values flagged (unique_IDs starting with 'P' and 'F1' kept)",
                  tr, length(extreme)))

    ## 3. save BEFORE plot
    p_before <- ggplot(dat, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: before"))

    ## 4. remove those extremes unless ID starts with P or F1
    dat2 <- dat %>%
      mutate(
        {{ tr }} := if_else(
          .data[[tr]] %in% extreme & !grepl("^(P|F1)", unique_ID),
          NA_real_,
          .data[[tr]]
        )
      )

    ## 5. AFTER plot
    p_after <- ggplot(dat2, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: after"))

    ## 6. sanity‑check: Rosner should find zero now
    if (k > 0) {
      
      # remove the parent and F1 observations from the data before testing
      vec_keep <- dat2 %>%     
      filter(!grepl("^(P|F1)", unique_ID)) %>%
      pull({{ tr }})
      
      remaining <- rosnerTest(vec_keep,
                            k = length(extreme))$n.outliers
      if (remaining > 0)
      warning(glue::glue("{tr}: {remaining} outlier(s) still present"))
    }
    
    ## 7. store updated data and plots
    acc$data  <- dat2
    acc$plots[[tr]] <- p_before | p_after
    acc
  })

flo_morph_by_ID <- result$data
plots    <- result$plots          # list of before|after panels

# plot each trait
plots[1]
plots[2]
plots[3]
plots[4]
plots[5]
plots[6]
plots[7]
plots[8]
plots[9]
plots[10]
plots[11]
plots[12]
plots[13]
plots[14]

```

# Nectar data
## Load and filter data
```{r load nectar}
nectar <- read_sheet(ss="19XHi3K57mDi2BpMPrlNOR2w16VSwV0YP6WYKYJHvuqs", sheet="nectar")

brix_density <- read.csv("~/Dropbox/Costus/costus-genetic-mapping/phenotype/data/brix_density_table.csv", header = TRUE)

#make a new column that makes a unique ID for each plant
nectar <- mutate(nectar, unique_ID = paste0(plant_type, "_", ID))

#exclude columns we don't need
nectar <- nectar %>% dplyr::select(-year, -date, -plant_type, -ID, -rep, -fl_capsize, -fl_nectarlength_mm, -fl_mg_sucrose)

# calculate mass of sucrose in solution
EFN_percent_sucrose_rounded <- round(nectar$EFN_percent_sucrose)

densities <- c()
for (i in 1:length(EFN_percent_sucrose_rounded)) {
  density <- brix_density$Density[brix_density$Brix == EFN_percent_sucrose_rounded[i]]
  densities <- c(densities, density)
}

EFN_H2Ovolume_mL = nectar$EFN_H2Ovolume_uL / 1000
mass_solution <- EFN_H2Ovolume_mL * densities
mass_sugar_grams <- (EFN_percent_sucrose_rounded/100) * mass_solution
nectar$mass_sugar_mg <- mass_sugar_grams * 1000

#exclude columns we don't need
nectar <- nectar %>% dplyr::select(-EFN_H2Ovolume_uL, -EFN_percent_sucrose)

#rename columns
nectar <- nectar %>% dplyr::rename("VFN" = "fl_nectarvolume_uL", # volume floral nectar
                                   "FNSC" = "fl_percent_sucrose", # percent sucrose floral nectar
                                   "EFNSC" = "mass_sugar_mg") # mass of sucrose in mg

#quick look at data
nectar
```

## Collapse the replicate observations
```{r nectar collapse reps}
nectar_by_ID <- nectar %>% group_by(unique_ID) %>% reframe(
  tibble(
    across(where(is.double), \(x) mean(x, na.rm = TRUE)),
    across(where(is.character), Modes),
    across(where(is.factor), Modes)
  )
)
```

## Identify potential outliers 
```{r potential outliers nectar}

traits <- setdiff(names(nectar_by_ID), c("unique_ID"))

result <- reduce(traits,           # loop over traits
  .init = list(data = nectar_by_ID, plots = list()),
  .f = function(acc, tr) {

    dat <- acc$data
    
    #vec <- dat[[tr]]
    
    vec <- dat %>%                       # rows that CAN be tested/removed
      filter(!grepl("^(P|F1)", unique_ID)) %>%
      pull({{ tr }})

    ## 1. Rosner test – how many to remove?
    if (length(boxplot.stats(vec)$out) == 0) {
      test.k <- 1
    } else {
      test.k <- length(boxplot.stats(vec)$out)
    }
    
    k <- rosnerTest(vec, k = test.k)$n.outliers

    ## 2. pick k single most‑extreme values
    outs    <- boxplot.stats(vec)$out
    center  <- median(vec, na.rm = TRUE)
    extreme <- outs[order(abs(outs - center), decreasing = TRUE)][seq_len(k)]
    
    message(sprintf("\n\nTrait %s – %d extreme values flagged (unique_IDs starting with 'P' and 'F1' kept)",
                  tr, length(extreme)))

    ## 3. save BEFORE plot
    p_before <- ggplot(dat, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: before"))

    ## 4. remove those extremes unless ID starts with P or F1
    dat2 <- dat %>%
      mutate(
        {{ tr }} := if_else(
          .data[[tr]] %in% extreme & !grepl("^(P|F1)", unique_ID),
          NA_real_,
          .data[[tr]]
        )
      )

    ## 5. AFTER plot
    p_after <- ggplot(dat2, aes(x = "", y = .data[[tr]])) +
      geom_boxplot(fill = "#0c4c8a") +
      theme_minimal() +
      ggtitle(glue::glue("{tr}: after"))

    ## 6. sanity‑check: Rosner should find zero now
    if (k > 0) {
      
      # remove the parent and F1 observations from the data before testing
      vec_keep <- dat2 %>%     
      filter(!grepl("^(P|F1)", unique_ID)) %>%
      pull({{ tr }})
      
      remaining <- rosnerTest(vec_keep,
                            k = length(extreme))$n.outliers
      if (remaining > 0)
      warning(glue::glue("{tr}: {remaining} outlier(s) still present"))
    }
    
    ## 7. store updated data and plots
    acc$data  <- dat2
    acc$plots[[tr]] <- p_before | p_after
    acc
  })

nectar_by_ID <- result$data
plots    <- result$plots          # list of before|after panels

# combine all plots into grid
wrap_plots(plots, nrow = length(traits))   # grid of all traits
```

# Combine and write data
```{r combine data}
data <- inflor_by_ID %>% 
  full_join(redarea_by_ID, by="unique_ID") %>%
  full_join(flo_morph_by_ID, by="unique_ID") %>%
  full_join(nectar_by_ID, by="unique_ID")


write.csv(data, "~/Dropbox/Costus/costus-genetic-mapping/phenotype/results/processed_data/phenotypic_data_no_outliers_with_parents_F1s.csv", quote = FALSE, row.names = FALSE)
```

## Exclude parents and F1s
```{r exclude parents and F1s}
#exclude parents and F1s
inflor_by_ID_filtered <- inflor_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

redarea_by_ID_filtered <- redarea_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

flo_morph_by_ID_filtered <- flo_morph_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

nectar_by_ID_filtered <- nectar_by_ID %>%
  filter(!grepl("^(F1|P)", unique_ID))

data <- inflor_by_ID_filtered %>% 
  full_join(redarea_by_ID_filtered, by="unique_ID") %>%
  full_join(flo_morph_by_ID_filtered, by="unique_ID") %>%
  full_join(nectar_by_ID_filtered, by="unique_ID")


write.csv(data, "~/Dropbox/Costus/costus-genetic-mapping/phenotype/results/processed_data/phenotypic_data_no_outliers_exclude_parents_F1s.csv", quote = FALSE, row.names = FALSE)

```
